---
title: "Diversity and Distribution course support"
format:
    html:
      number-sections: true
      toc: true
editor: visual
# https://quarto.org/docs/computations/execution-options.html#output-options
execute:
  # show code chunk output
  include: true
  # Show the code in the output
  echo: true
  # Show messages
  message: true
  # Show warnings
  warning: true
  # Cache code results
  cache: true
bibliography: references.bib
---

```{r}
#| label: Install
#| include: false
### Utilities for R.

# Necessary packages
packages_required <- c(
  # R tools
  "knitr", "tidyverse", 
  # Scientific packages
  "ade4", "divent", "GoFKernel", "sads", "untb", "vegan"
)
# Already installed?
packages_installed <- vapply(
  packages_required, 
  FUN = requireNamespace, 
  FUN.VALUE = TRUE, 
  quietly = TRUE
) 
# Install missing packages
install.packages(names(packages_installed)[!packages_installed])

# Chunk font size hook: allows size='small' 
# or any valid Latex font size in chunk options
def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(
  chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(
      options$size == "normalsize", 
      yes = x,
      no = paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize")
    )
  }
)
```

# Motivation

This document contains code chunks to help you understanding the course. Don't hesitate to modify them, add your own notes and code.

It comes with a practical you will run at home (`Biodiv-Practical.Rmd`).

# Species distributions

## Figures

### Preston diagram

Using the vegan package, we can access the BCI data. The data is 50 rows representing 50 plots, and each species with a number of individuals \>10cm of that species per plot. For our purposes we will combine all the plots to obtain species totals over 50 hectares (with a total number of trees = 21,457, and per hectare = 429). There are 600 trees/ha in Paracou (also they're smaller). The distribution is Gaussian (normal) and must be logged cause there is sooo many trees with very few individuals.

```{r}
library("tidyverse")
library("vegan")
data(BCI)

# Sum the 50 1-ha plots. this is a vector
BCI_50ha_abd <- colSums(BCI)

#make the Preston histogram by taking the log and breaking into 15 equal chunks
BCI_50ha_abd %>% 
  log10() %>% 
  hist(breaks = 15, main = "Preston diagram")
```

Probability densities

Ok then we make a smoothed version to more clearly see the pattern vs the histogram. This is not technically a Preston diagram - it's a probablility density

```{r}
BCI_50ha_abd %>% 
  tibble(abundance = .) %>% #this changes our vector to a tibble of column name abundance -> simple df. basic pipeline would crash here!!!
  ggplot() +
  # Bounded density - bounded from 0 to infinity, so that nothing is <0. cause there are no trees with 0 individuals. log(0) -> 1
  geom_density(aes(x = abundance), bounds = c(0, Inf)) +
  scale_x_log10()
```

### Whittaker diagram

Preferred diagram! Shows number of species of different types of abundances

```{r}
library("divent")
BCI_50ha_abd %>% 
  # Declare it is an "abundances" object
  as_abundances() %>% 
  autoplot()
```

## Broken stick

Random distributions are created by `divent::rxxx()`functions. Rcommunity is used to random communities of individuals

```{r}
rcommunity(
  # One community
  1,
  # Number of trees
  size = 10000,
  # Number of species
  species_number = 300,
  # Distribution type - broken stick
  distribution = "bstick"
) %>% 
  # Whittaker plot
  autoplot()
```

## Geometric

Random distribution. Only shows to \~70 species cause all the others have less than 1 (so 0) individuals out of 10,000. Increasing the number of trees would show us more species. Though it should theoretically be a straight line, it is not because of integers. Not used very often cause you need A LOT of individuals. Used most often for soil microbes.

```{r}
rcommunity(
  # One community
  1,
  # Number of trees
  size = 10000,
  # Number of species
  species_number = 300,
  # Distribution type - geometric
  distribution = "geom",
  prob_geom = 0.1 #choosing 10% as amount of resources 1st species recieves
) %>% 
  # Whittaker plot
  autoplot()
```

## Lognormal

Random distribution.

```{r}
rcommunity(
  # One community
  1,
  # Number of trees
  size = 10000,
  # Number of species
  species_number = 300,
  # Distribution type - lognormal
  distribution = "lnorm",
  sd_lnorm = 1
) %>% 
  # Whittaker plot
  autoplot()
```

Modify the standard deviation.

## Log-series

The gamma function, which shows lots of species with few individuals, some species with a few more, and very few species with many individuals. K = 1 (shape = 1) is what makes it gamma. According to Fisher, it should be very close to 0.

```{r}
ggplot() +
  geom_function(
    fun = dgamma, 
    args = list(shape = 0.1),
    xlim =c(0,10)
  )
```

Random distribution. Used for metacommunities

```{r}
rcommunity(
  # One community
  1,
  # Number of trees = n
  size = 10000,
  # Number of species is calculated = S
  # Distribution type - log series
  distribution = "lseries",
  fisher_alpha = 40 #this is the value at BCI
) %>% 
  # Whittaker plot
  autoplot()
```

## Estimate the parameters

Draw a very large lognormal distribution.

```{r}
rcommunity(
  # One community
  1,
  # Number of trees = max supported integer value
  size = .Machine$integer.max,
  # Number of species
  species_number = 300,
  # Distribution type - log normal
  distribution = "lnorm",
  sd_lnorm = 1
) %>% 
  # Name it
  mutate(site = "Large log-normal") ->
  comm_large_lnorm

# Plot it
comm_large_lnorm %>% 
  autoplot()
```

Sample it. Basically, denote something as a local community. To use the abundance, use abd and insert a vector.

```{r}
# Draw 3000 trees from the large community
rcommunity(
  1,
  size = 3000,
  abd = as.numeric(comm_large_lnorm) #drawn from the abundances of the metacommunity
) %>% 
  # Name it
  mutate(site = "Small log-normal") ->
  comm_small_lnorm

# Plot it
comm_small_lnorm %>% 
  autoplot()

# Make a single tibble with all data
comm_large_lnorm %>% 
  # Make a single tibble
  bind_rows(comm_small_lnorm) ->
  comm_lnorm

# Plot the distributions and fit a lognormal model. fit-rac makes a line
autoplot(comm_lnorm, fit_rac = TRUE, distribution = "lnorm")

# Estimate the parameters
fit_rac(comm_lnorm, distribution = "lnorm") 
```

BCI example. Doesn't fit as well as simulated data, cause simulated data always fits more exactly than observed data. Can try it with different types of distributions to see which one fits best (hint, it's log normal).

```{r}
BCI_50ha_abd %>% 
  as_abundances() %>% 
  autoplot(fit_rac = TRUE, distribution = "lnorm") # can try lseries, bstick, geom

```

## Zero-sum multinomial distribution

BCI parameters $\theta$ and $m$ are estimated following @etienne2005 with the untb package. This code chunk has a lot of bugs and issues cause the data are toooo big. still runable, by bypassing the bugs and just re-downloading the data. Remember that $\theta$ is equal to $\alpha$ !

```{r}
#| label: bci_params
#| echo: true
# BCI data
library("vegan")
data(BCI)
BCI_50ha <- colSums(BCI)
# Don't load untb because of conflicts with dplyr::count() and select()
# library("untb")
if (file.exists("data/BCI_logkda.RData")) {
  # Use the saved result if available to save
  # computation resources
  load("data/BCI_logkda.RData")
} else {
  BCI_50ha %>% 
  # name conflict with dplyr::count()
  untb::count() %>% 
  # requires pari/gp installed and in the search paths,
  # or use gp_binary argument
  untb::logkda() ->
  BCI_logkda
  # Save the result
  save(BCI_logkda, file = "data/BCI_logkda.RData")
}
BCI_50ha %>% 
  # name conflict with dplyr::count()
  untb::count() %>% 
  # Estimate theta and m
  untb::optimal.params(log.kda = BCI_logkda) %>% 
  print() ->
  BCI_params
```

The distribution of the local community is derived. $J$ is that of BCI and $m$ was estimated previously. Here we run a simulated version of the local BCI data, based on the metacommunity BCI data, with a line on top as the observed data. This is just a longer by hand version of simulating a distribution that exactly fits the neutral model. Can instead just run the previous plain BCI log normal distribution

```{r}
#| label: bci_sim_code
#| echo: true
J <- sum(BCI) #local community
theta <- BCI_params[1] #metacommunity 
m <- BCI_params[2] #local community, but same as from metacommunity
library("sads")
# Number of species in the local community
Svolkov(theta, m, J) %>% 
  # Must be integers for the simulation
  round() %>% 
  # Draw the species abundances - very computer intense, beware with large datasets
  rvolkov(theta, m, J) %>% 
  # Make it an abundances object to plot it
  as_abundances() %>% 
  # Whittaker plot
  autoplot() ->
  BCI_sim_neutral

# BCI real distribution
BCI_50ha %>% 
  sort(decreasing = TRUE) %>% 
  as_tibble() %>% 
  mutate(rank = seq_len(n())) -> 
  BCI_tbl_50ha

# Plot altogether
BCI_sim_neutral +
  geom_line(
    data = BCI_tbl_50ha,
    aes(x = rank, y = value)
  )
```

# Species Accumulation Curves

## BCI data

Gather all subplots.

```{r}
library("vegan")
data(BCI)
BCI %>% 
  # Make an abundances object
  as_abundances() %>% 
  # Make all plot weights equals
  mutate(weight = 1) %>% 
  # Gather all 50 plots
  metacommunity() %>% 
  # Name it
  mutate(site = "BCI") -> 
  BCI_all
autoplot(BCI_all, fit_rac = TRUE, distribution = "lnorm")
fit_rac(BCI_all, distribution = "lnorm") 
```

Abundance frequency counts help read the figure.

```{r}
BCI %>% 
colSums() %>% 
abd_freq_count() %>% 
arrange(desc(abundance))
```

## Accumulation of subplots

Accumulate subplots in random order (Monte-Carlo). Plot the expectation and the confidence interval of the number of species with respect to the number of samples.

The function specaccum is a function from package vegan. It calculates row by row (plots) not by tree. Eric doesn't seem to like this much. prefers the other function.

```{r}
BCI %>% 
  specaccum() %>% 
  plot()
```

## Accumulation curve

Compute the expected accumulation curve of BCI up to twice the sampling effort, with its confidence interval.

Calculate the number of trees in BCI - simple and veeeery useful. Then we use the div_richness to estimate how many species we should have per this many trees. Then we make the smoothed curve for up to double our sample size, and we're only choosing every 10th trees, to cut the comp time. We only run 10 simulations, so it's not very precise. We would need to run 1000 simulations to have a good estimation

The graph: vertical line shows true number of individuals observed. confidence interval narrows further along the end. smoothing is not very clean, needs fixed. Shows what the kind of ideal sample size is in order for the data to be reliable.

```{r}
# Number of trees in BCI
BCI_n <- abd_sum(BCI_all, as_numeric = TRUE)
# Asymptotic richness
BCI_richness <- div_richness(BCI_all, as_numeric = TRUE)
BCI_all %>% 
  accum_hill(
    levels = seq(from = 1, to = 2 * BCI_n, by = 10),
    # Run simulations to estimate uncertainty.
    # Start with 10 simulations, run 100 later
    n_simulations = 10,
  ) %>% 
  autoplot() +
  geom_vline(xintercept = BCI_n, linetype = 2) +
  geom_hline(yintercept = BCI_richness, linetype = 1)
```

# Canonical lognormal

ggplot seems to bug with weighted, bounded density and log scale.

```{r}
# Most adundant species
BCI_50ha_mas <- max(BCI_50ha_abd)
BCI_50ha_abd %>% 
  tibble(abundance = .) %>% 
  ggplot() +
  # Bounded density
  geom_density(
    aes(x = abundance, color = "Species curve"), 
    bounds = c(0, Inf)
  ) +
  # Weighted density
  geom_density(
    aes(x = abundance, weight = abundance, color = "Individuals curve"), 
    bounds = c(min(BCI_50ha_abd), max(BCI_50ha_abd))
  ) +
  scale_x_log10()
```

Correct curves.

```{r}
library("GoFKernel")
plot(
  density.reflected(log10(BCI_50ha_abd), lower = 0),
  type = "l", 
  col = "red", 
  ylim = c(0, .8),
  main = "Canonical BCI",
  xlab = "log10(abundance)"
)
lines(
  density.reflected(
    log10(BCI_50ha_abd), 
    weights = BCI_50ha_abd, 
    lower = min(log10(BCI_50ha_abd)), 
    upper = max(log10(BCI_50ha_abd))
  )
)
abline(v = max(log10(BCI_50ha_abd)), lty = 2)

```

# Richness

## Dummy communities

Build dummy communities.

```{r}
tibble(
  sp_1 = c(10, 10, 1000, 1000),
  sp_2 = c(10, 11,    1,    0)
) %>% 
  as_abundances() %>% 
  print() ->
  comm_dummy
```

Sample richness.

```{r}
div_richness(comm_dummy, estimator = "naive")
```

Asymptotic richness.

```{r}
div_richness(comm_dummy)
```

## Paracou 6

Prepare the data: gather the four subplots.

```{r}
paracou_6_abd %>% 
  # Gather all sites
  metacommunity() %>% 
  # Rename "matacommunity"
  mutate(site = "Paracou 6") ->
  p6_abundances
autoplot(p6_abundances)
```

Abundance frequencies.

```{r}
p6_abundances %>% 
  as.numeric() %>% 
  abd_freq_count()
```

Estimators of richness.

```{r}
div_richness(p6_abundances, estimator = "naive") %>% 
  bind_rows(div_richness(p6_abundances, estimator = "Chao1")) %>% 
  bind_rows(div_richness(p6_abundances, estimator = "iChao")) %>% 
  # Default estimator
  bind_rows(div_richness(p6_abundances, estimator = "jackknife")) %>% 
  bind_rows(div_richness(p6_abundances, estimator = "jackknife", jack_max = 1))
```

## Interpolation / extrapolation

Diversity of 1 ha in Paracou.

```{r}
# Number of trees per hectare
p6_density <- abd_sum(p6_abundances, as_numeric = TRUE) / 6.25
# Richness
p6_abundances %>% 
  div_richness(level = round(p6_density))
```

Diversity of 10 ha in Paracou.

```{r}
# Richness
p6_abundances %>% 
  div_richness(level = round(10 * p6_density))
```

# Shannon

## Information function

Information brought by an individual: $I(p_s) = \ln(1 / p_s)$.

```{r}
ggplot() +
  # Increase the default number of points (101 to 1001) for precision
  geom_function(fun = \(p) {log(1 / p)}, n = 1001) +
  labs(x = "Probability", y = "Information")
```

Information brought by a species: $p_s \ln(1 / p_s)$.

```{r}
ggplot() +
  geom_function(fun = \(p) {p * log(1 / p)}, n = 1001) +
  labs(x = "Probability", y = "Information")
```

## Paracou 6

Estimators.

```{r}
ent_shannon(p6_abundances, estimator = "naive") %>% 
  bind_rows(ent_shannon(p6_abundances, estimator = "UnveilJ")) %>% 
  bind_rows(ent_shannon(p6_abundances, estimator = "ChaoJost")) %>% 
  print() ->
  p6_shannon
```

Effective number of species: $e^H$.

```{r}
p6_shannon %>% 
  mutate(diversity = exp(entropy))
```

## Dummy communities

Compare entropy.

```{r}
comm_dummy
ent_shannon(comm_dummy, estimator = "naive")
```

# Simpson

## Paracou 6

Estimators.

```{r}
ent_simpson(p6_abundances, estimator = "naive") %>% 
  bind_rows(ent_simpson(p6_abundances, estimator = "Lande")) %>% 
  bind_rows(ent_simpson(p6_abundances, estimator = "UnveilJ")) %>% 
  bind_rows(ent_simpson(p6_abundances, estimator = "ChaoJost")) %>% 
  print() ->
  p6_simpson
```

Effective number of species: $1 / (1 - E)$.

```{r}
p6_simpson %>% 
  mutate(diversity = 1 / (1 - entropy))
```

# Generalized entropy

## Information

Information brought by an individual: $I(p_s) = \ln_q(1 / p_s)$.

```{r}
ggplot() +
  # Increase the default number of points (101 to 1001) for precision
  geom_function(fun = \(p) {ln_q(1 / p, q = 0)}, n = 1001, aes(color = "0")) + 
  geom_function(fun = \(p) {ln_q(1 / p, q = 1)}, n = 1001, aes(color = "1")) + 
  geom_function(fun = \(p) {ln_q(1 / p, q = 2)}, n = 1001, aes(color = "2")) +
  scale_y_continuous(limits = c(0, 10)) +
  labs(x = "Probability", y = "Information", color = "q")
```

Information brought by a species: $p_s \ln_q(1 / p_s)$.

```{r}
ggplot() +
  # Increase the default number of points (101 to 1001) for precision
  geom_function(fun = \(p) {p * ln_q(1 / p, q = 0)}, n = 1001, aes(color = "0")) + 
  geom_function(fun = \(p) {p * ln_q(1 / p, q = 1)}, n = 1001, aes(color = "1")) + 
  geom_function(fun = \(p) {p * ln_q(1 / p, q = 2)}, n = 1001, aes(color = "2")) +
  labs(x = "Probability", y = "Information", color  = "q")
```

## Paracou 6

Diversity profile.

```{r}
p6_abundances %>% 
  profile_hill() %>% 
  autoplot() +
  # Don't print the legend
  theme(legend.position = "none")
```

Compare estimators.

```{r}
p6_abundances %>% 
  profile_hill(estimator = "UnveilJ") %>%
  # Use the site name to store the estimator's name
  mutate(site = "UnveilJ") %>% 
  bind_rows(
    p6_abundances %>% 
    profile_hill(estimator = "ChaoJost") %>% 
    mutate(site = "ChaoJost")
  ) %>% 
  bind_rows(
    p6_abundances %>% 
    profile_hill(estimator = "naive") %>% 
    mutate(site = "naive")
  ) %>% 
  autoplot() +
  labs(color  = "Estimator")
```

Extreme values.

The reciprocal of the Berger-Parker index [@Berger1970] is the rarity of the most abundant species $^{+\infty}D$.

```{r}
# q = 1000 is enough for infinity
div_hill(p6_abundances, q = 1000)
# Compare to 
p6_abundances %>% 
  # Extract proobabilities
  as_probabilities(as_numeric = TRUE) %>% 
  # Max value
  max() %>% 
  # Reciprocal
  `^`(-1)
  
```

## Dummy communities

Compare diversity

```{r}
comm_dummy
div_hill(comm_dummy, q = 1, estimator = "naive")
```

# Comparing diversity

## Asymptotic diversity

Compare asymptotic diversity of Paracou and BCI, extend $q$.

```{r}
# Orders from 0 to 4
q_seq_0_4 <- seq(from = 0, to = 4, by = 0.1)
profile_hill(p6_abundances, orders = q_seq_0_4) %>% 
  bind_rows(profile_hill(BCI_all, orders = q_seq_0_4)) %>% 
  autoplot()
```

## Standardized diversity

The sampling effort is very different: 50 ha in BCI, 6.25 ha in Paracou. The asymptotic estimation is less reliable in Paracou.

Compare diversity at the same sampling effort, say 10 ha.

```{r}
# Number of trees in 10 ha
p6_n_10ha <- round(abd_sum(p6_abundances, as_numeric = TRUE) / 6.25 * 10)
BCI_n_10_ha <- round(sum(BCI_all) / 5)
```

Compare diversity of 10 ha.

```{r}
profile_hill(p6_abundances, level = p6_n_10ha) %>% 
  bind_rows(profile_hill(BCI_all, level = BCI_n_10_ha)) %>% 
  autoplot()
```

## Sample coverage

```{r}
coverage(p6_abundances) %>% 
  bind_rows(coverage(BCI_all))
```

Compare diversity at the same sample coverage [@Chao2012b], that of Paracou.

```{r}
reference_coverage <- coverage(p6_abundances, as_numeric = TRUE)
profile_hill(p6_abundances, level = reference_coverage) %>% 
  bind_rows(profile_hill(BCI_all, level = reference_coverage)) %>% 
  autoplot()
```

# Phylogenetic diversity

## Mock data

A simple community with 3 species.

```{r}
mock_3sp_abd %>% 
  as_abundances() %>% 
  mutate(site = "Mock 3 species") %>% 
  print() ->
  mock_3sp_abundances
```

```{r}
library("ape")
# Prepare the tree for convenience: make the branches twice longer
mock_3sp_tree$edge.length <- 2 * mock_3sp_tree$edge.length
plot(mock_3sp_tree, direction = "downwards", srt = 90)
axis(2)
```

PD is the total length of branches

```{r}
div_pd(mock_3sp_abd, tree = mock_3sp_tree)
```

Rao.

```{r}
ent_rao(mock_3sp_abundances, tree = mock_3sp_tree, estimator = "naive")
```

The tree is normalized to height = 1 by default. Without normalization:

```{r}
ent_rao(
  mock_3sp_abundances, 
  tree = mock_3sp_tree, 
  estimator = "naive", 
  normalize = FALSE
)
```

## Paracou 6

`paracou_6_taxo` is a taxonomic tree: distance is 1 between species of the same genus, 2 between genera of the same family, 3 between families.

```{r}
plot(paracou_6_taxo, type = "fan", show.tip.label = FALSE) 
```

Calculate Rao's entropy and phylogenetic entropy of order 2

```{r}
ent_rao(p6_abundances, tree = paracou_6_taxo) %>% 
  bind_rows(
    ent_phylo(p6_abundances, tree = paracou_6_taxo, q = 2)
  )
```

Rao's entropy is not an effective number of species, but the average phylogenetic distance between two individuals. The corresponding Hill number is:

```{r}
div_phylo(p6_abundances, tree = paracou_6_taxo, q = 2)
```

Diversity profile

```{r}
profile_phylo(p6_abundances, tree = paracou_6_taxo) %>% 
  autoplot()
```

# Functional diversity

## Mock data

The same mock data is used. Distances between species is now assumed to be in the trait space. Distances between species.

```{r}
mock_3sp_dist
```

Plot the trait space.

```{r}
library("ade4")
mock_pcoa <- dudi.pco(mock_3sp_dist, scannf = FALSE)
scatter(mock_pcoa, posieig = "none")
```

Similarity between species.

```{r}
# The dist object must be converted to a matrix
mock_3sp_Z <- 1 - as.matrix(mock_3sp_dist) / max(mock_3sp_dist)
mock_3sp_Z
```

Ordinariness.

```{r}
# Probabilities of species
mock_3sp_prob <- mock_3sp_abd / sum(mock_3sp_abd)
# Ordinariness of species
mock_3sp_O <- as.numeric(mock_3sp_Z %*% mock_3sp_prob)
# Compare probabilities and ordinariness
tibble(
  Probability = mock_3sp_prob,
  Rarity_p = 1 / mock_3sp_prob,
  Ordinariness = mock_3sp_O,
  Rarity_0 = 1 / mock_3sp_O
)
```

Diversity.

```{r}
profile_similarity(
  mock_3sp_abundances, 
  similarities = mock_3sp_Z, 
  estimator = "naive"
) %>% 
  autoplot()
```

Diversity of order 2 (i.e. Rao's).

```{r}
ent_similarity(
  mock_3sp_abundances, 
  similarities = mock_3sp_Z, 
  q = 2, 
  estimator = "naive"
)
```

## Paracou 6

The species traits are maximum height (95^th^ percentile) and seed mass, maximum height (95^th^ percentile) and seed mass. A distance matrix was built (see home practical). Distribution of distances between species.

```{r}
paracou_6_fundist %>% 
  as.numeric() %>% 
  tibble(Distance = .) %>% 
  ggplot() +
  geom_density(aes(x = Distance))
```

Diversity profile.

```{r}
p6_abundances %>% 
  profile_similarity(
    similarities = 1 - as.matrix(paracou_6_fundist) / max(paracou_6_fundist)
  ) %>% 
  autoplot()
```

## Intraspecific variability

Coordinates of species A, B and C in the trait space.

```{r}
mock_pcoa$li
```

Plot individuals in the trait space, with a small variability.

```{r}
# Create coordinates
tibble(
  Species = rep(names(mock_3sp_abd), times = mock_3sp_abd),
  x = rep(mock_pcoa$li[, 1], times = mock_3sp_abd),
  y = rep(mock_pcoa$li[, 2], times = mock_3sp_abd)
) ->
  mock_3sp_ind
# Choose variability
variability <- 0.1
# Map
mock_3sp_ind %>% 
  # Add variability
  mutate(
    x = x + rnorm(n(), mean = 0, sd = variability),
    y = y + rnorm(n(), mean = 0, sd = variability)
  ) %>% 
  ggplot(aes(x = x, y = y, color = Species)) +
  geom_point() +
  coord_fixed()
```

Diversity of the 6 individuals. First, without variability.

```{r}
# Choose variability
variability <- 0
# Create the point set
mock_3sp_ind %>% 
  # Add variability
  mutate(
    x = x + rnorm(n(), mean = 0, sd = variability),
    y = y + rnorm(n(), mean = 0, sd = variability)
  )  %>% 
  # Distance matrix
  dist() %>% 
  as.matrix() ->
  mock_6ind_dist
# Delete the row and column name of the matrix (they are not species names)
rownames(mock_6ind_dist) <- colnames(mock_6ind_dist) <- NULL
# Total abundance
mock_6ind_n <- nrow(mock_6ind_dist)
# Similarity
mock_6ind_Z <- 1 - mock_6ind_dist / max(mock_6ind_dist)
# Abundances of individuals are 1, or any identical value
mock_6ind_abd <- rep(1, mock_6ind_n)
# Diversity
profile_similarity(
  mock_6ind_abd, 
  similarities = mock_6ind_Z,
  estimator = "naive"
) %>% 
  autoplot()
```

Change the variability (0.1 first) and run the code again.

# Diversity partitioning

## Mock data

Three habitats: A, B, C.

```{r}
matrix(
  c(
    0, 0, 0, 0, 0, 1, 1, 1, 1, 1,
    1, 1, 1, 1, 1, 1, 0, 1, 0, 0,
    0, 0, 0, 0, 0, 0, 0, 1, 1, 1
  ),
  nrow = 3, 
  byrow = TRUE
) %>% 
  as_abundances() %>% 
  mutate(
    site = paste("Habitat", LETTERS[1:3]),
    weight = 1
  ) ->
  mock_island_abd
```

Richness.

```{r}
div_richness(mock_island_abd, estimator = "naive")
div_richness(mock_island_abd, gamma = TRUE, estimator = "naive")
```

Multiplicative partitioning: 10 effective species ($\gamma$) = 5 effective species per habitat ($\alpha$) times 2 habitats ($\beta$).

```{r}
div_part(mock_island_abd, q = 0, estimator = "naive")
```

Additive partitioning: 10 effective species ($\gamma$) = 5 effective species per habitat ($\alpha$) + 5 more species ($\beta$).

## Paracou 6

The plot is made of 4 subplots.

```{r}
# Diversity profiles of the 4 subplots
paracou_6_abd %>% 
  profile_hill() %>% 
  autoplot()
```

### Partitioning

Function `div_part()` computes $\alpha$, $\beta$ and $\gamma$ diversity.

```{r}
div_part(paracou_6_abd, q = 1)
```

### Step by step

```{r}
# Alpha entropy
paracou_6_abd %>% 
  ent_tsallis(q = 1) %>% 
  # Mean entropy
  group_by(order) %>% 
  summarise(ent_alpha = mean(entropy)) %>% 
  # Diversity
  mutate(div_alpha = exp_q(ent_alpha, q = 1)) %>% 
  print() ->
  paracou_6_alpha
```

```{r}
# Gamma entropy
paracou_6_abd %>% 
  # Group the subplots
  metacommunity() %>% 
  ent_tsallis(q = 1) %>%
  # Rename entropy for clarity
  rename(ent_gamma = entropy) %>% 
  # Diversity
  mutate(div_gamma = exp_q(ent_gamma, q = 1)) %>% 
  print() ->
  paracou_6_gamma
```

```{r}
# Make a single tibble with alpha and gamma diversities
paracou_6_alpha %>% 
  bind_cols(
    paracou_6_gamma %>% 
      select(ent_gamma, div_gamma)
  ) %>% 
  mutate(
    # beta entropy
    ent_beta = ent_gamma - ent_alpha,
    # beta diversity
    div_beta = div_gamma / div_alpha
  )
```

### Profiles

Not implemented yet in divent.

alpha diversity.

```{r}
# Values of q used for computation: default values of profile_xxx()
orders <- seq(from = 0, to = 2, by = 0.1)
sapply(
  # Apply ent_tsallis to all orders
  orders,
  FUN = \(q) {
    # Get a number to obtain a matrix with sapply
    ent_tsallis(paracou_6_abd, q = q, as_numeric = TRUE)
  }
) %>% 
  # Each row contains a subbplot, each column an order, values are entropy
  # Calculate mean entropy (weights are equal)
  colMeans() %>% 
  # Diversity
  exp_q(q = orders) %>% 
  # Make a tibble
  tibble(
    order = orders,
    div_alpha = .
  ) %>% 
  print() ->
  paracou_6_alpha_profile
```

gamma diversity.

```{r}
paracou_6_abd %>% 
  profile_hill(gamma = TRUE) %>% 
  # Rename diversity to recongnize gamma later
  rename(div_gamma = diversity) %>% 
  print() ->
  paracou_6_gamma_profile
```

Compute beta.

```{r}
paracou_6_gamma_profile %>% 
  # Join the tibbles by identical order
  inner_join(paracou_6_alpha_profile) %>% 
  # Calculate beta diversity
  mutate(div_beta = div_gamma / div_alpha) %>% 
  print() ->
  paracou_6_profile
```

Plot.

```{r}
paracou_6_profile %>% 
  # paracou_6_profile is an object of class profile that can be plotted
  # The column to plot must be named diversity
  rename(diversity = div_alpha) %>% 
  autoplot() + 
  theme(legend.position = "none")
```

```{r}
paracou_6_profile %>% 
  # rename div_beta to plot it
  rename(diversity = div_beta) %>% 
  autoplot() + 
  theme(legend.position = "none")
```
